Pipit: Reactive Systems in \fstar ~(Extended Abstract)

### 2023-05-10 2023pipit/Main.tex
intro page 1

\begin{abstract}
  Safety-critical control systems, such as the anti-lock braking systems in most cars today, need to be correct and execute in real-time.
  We could write these systems directly in an imperative language such as C or Rust and verify them, but it can be difficult to state high-level specifications about such low-level implementations.
  Another approach, favoured by large parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre or SCADE, and verify that the implementations satisfy the high-level specification using a model-checker.
  These model-checkers can prove many interesting properties automatically, but do not provide many options when the automated proof techniques fail.
  We introduce Pipit, an embedded domain-specific language for implementing and verifying such controllers in \fstar.
  Pipit aims to provide a high-level language similar to Lustre, while reusing \fstar's proof automation and manual proofs for verifying controllers, and using LowStar's C-code generation for real-time execution.
  Pipit translates its expression language to a transition system for k-inductive proofs, which is mostly verified; verifying the translation to imperative code is future work.
\end{abstract}

@@

@@

@@

>>>>

\begin{abstract}
  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
  We introduce Pipit, a simple reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved using k-induction on the transition system.
  Pipit can also generate imperative code in \lowstar{}, a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
  This translation to imperative code has been proved to preserve types; the proof that the generated imperative code preserves semantics is future work.
\end{abstract}

@@

Safety-critical control systems, such as the anti-lock braking systems that occur in most cars today, need to be correct and execute in real-time.
One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
This mismatch may occur even when the compiler has been verified correct, as in the case of Vélus~\cite{bourke2017formally}.
For example, in the Vélus compiler, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB}.

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In the following, we introduce Pipit, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on a clock-free subset of Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers, and using \lowstar{}'s C-code generation for real-time execution.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
Pipit translates its expression language to a transition system for k-inductive proofs, which is verified\footnote{The translation is verified for a previous untyped expression representation; we are still finishing porting the proofs to the new typed representation.}; verifying the translation to imperative code is future work.

Our contributions are as follows:
\begin{itemize}
  \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
  \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
    % TODO: CUT WAFFLE
    This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
    Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
  \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
\end{itemize}

% \begin{itemize}
%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
% \end{itemize}

\section{Pipit in flight}

\begin{tabbing}
  @MM@\= @MMMMMM@ \= \kill
  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
    \> @let rec@ $p'$ $= (\xfby{\top}{p'}) \wedge p$ \\
    \> @in @ $p'$
\end{tabbing}

  \begin{tabbing}
    @MM@\= @MMMMMM@ \= \kill
    @let@ fir (coefficients: list $\RRgez$) (s: stream $\RR$): stream $\RR$ = \\
      \> @match@ coefficients @with@ \\
      \> @|@ $[]$ \> $\to 0$ \\
      \> @|@ $c :: \cs $ \> $\to (s \cdot c) + \mbox{fir~} \cs ~(\xfby{0}{s})$ \\
  \end{tabbing}

$\blacksquare(s = 0) \implies \mbox{fir~} \cs~s = 0 \blacksquare \Box \blacklozenge \Diamond$

@@

\section{Coffee machine}
\label{s:coffee-machine}

@@

\TODO{discuss TCB: big, much bigger than Vélus}
